---
title: "Disease Dynamics and Modelling Assignment 2"
subtitle: Solving the SEIR model
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
  word_document: default
---

# Walkthrough: how to numerically solve an SEIR model

A novel pathogen has been discovered in 3 individuals in a population of 1000 individuals, with 2 individuals exposed but not infectious and 1 individual exposed and infectious.
It has been found to have an incubation/progression period of 5 days, an infectious period of 10 days and a daily transmission rate of 1.7. 
You are to analyse the spread of this pathogen during the first 100 days of the outbreak using an SEIR model.
Find and plot the solution.

Step 1: write a function that computes the system of ODEs and returns a list containing the derivatives. The input parameters of this function are time `t`, the compartments as a vector `x`, and model parameters `parameters`. The outputs of the function are a list of derivatives. We name this function as `SEIR.model`.
```{r}
SEIR.model <- function(t, x, parameters){ 
  # Get the current number of individuals in each compartment.
  S <- x[1]
  E <- x[2]
  I <- x[3]
  R <- x[4]
  
  # Get the model parameters.
  beta <- parameters$beta
  rho <- parameters$rho
  gamma <- parameters$gamma
  
  # Calculate the total number of individuals in the system.
  # This is just all the compartments added together.
  N <- S + E + I + R
  
  # Calculate the derivatives (the system of ODEs).
  dS <- -beta*S*I/N
  dE <- beta*S*I/N - rho*E
  dI <- rho*E - gamma*I
  dR <- gamma*I
  
  # Return the derivatives inside a list.
  derivatives <- list(c(dS, dE, dI, dR))
  return(derivatives)
}
```

Step 2: define the initial conditions, parameters and times at which you wish to compute the solution.
```{r}
# The initial parameters: how many people are in each compartment at the start of
# the model?
init <- c(S = 997, E = 2, I = 1, R = 0.0)

# The model parameters: what is the transmission rate (beta), progression rate (rho) and recovery rate (gamma).
pars <- list(beta = 1.7, rho = 0.2, gamma = 0.1)

# The solution times: the sequence of times for which to compute the solution.
time <- seq(from = 0, to = 100, by = 1)
```

Step 3: solve the system using the 'lsoda' function from the 'deSolve' package (try help(lsoda) or ??Isoda for help). 
## ode vs lsoda
```{r}
# Load in the deSolve package.
library(deSolve)

# lsoda is a function with inputs: ODE system (as a function), initial conditions,
# parameters and times to compute the solution. The solution for the SIR model is
# stored in 'out'.
out <- lsoda(func = SEIR.model, y = init, parms = pars, times = time)

# Convert the solution into a data.frame. This makes plotting easier.
out <- as.data.frame(out)
head(out)
```

Step 4: plot the solution for all compartments on a single plot.
```{r}
# First plot time (x-axis) against the number of susceptible (y-axis).
plot(out$time, out$S,
     type = 'l', col = 'green', ylab = "Number of individuals in each compartment",
     xlab = "Time (days)", lwd = 3, ylim = c(0, 1000), xlim = c(0,103))

# Next, we can add lines for time against infecteds, and time against recovered/removed.
lines(out$time, out$E, col = 'blue', lwd = 3)
lines(out$time, out$I, col = 'red', lwd = 3)
lines(out$time, out$R, col = 'grey', lwd = 3)

# Finally, we can add a legend so we know what lines correspond to what compartment.
legend("topright", legend = c("S","E","I","R"), col = c("green","blue","red","grey"), lwd = 2)
```
```{r}
# view the last days of the model
tail(round(out))
```


# Exercises: exploring the SEIR model
Now we can make some modifications to the model parameters, and observe how the model output changes.
```{r}
myplotter <- function(out){
  # First plot time (x-axis) against the number of susceptible (y-axis).
  plot(out$time, out$S,
     type = 'l', col = 'green', ylab = "Number of individuals in each compartment",
     xlab = "Time (days)", lwd = 3, ylim = c(0, 1000), xlim = c(0,103))

  # Next, we can add lines for time against infecteds, and time against recovered/removed.
  lines(out$time, out$E, col = 'blue', lwd = 3)
  lines(out$time, out$I, col = 'red', lwd = 3)
  lines(out$time, out$R, col = 'grey', lwd = 3)

  # Finally, we can add a legend so we know what lines correspond to what compartment.
  legend("topright", legend = c("S","E","I","R"), col = c("green","blue","red","grey"), lwd = 2)
}
```

## Exercise 1

Add vital dynamics to the model i.e. birth and death, assuming;
i) All individuals can reproduce regardless of the compartment
ii) All newborn individuals are born into the susceptible compartment
iii) The population is constant i.e. birth and death rate are equal

Given the same population and parameters, with the daily birth/death rate (mu) = 0.0002, plot the trajectory of all the compartments in 100 days.

```{r}
modified.SEIR.model <- function(t, x, parameters){ 
  # Get the current number of individuals in each compartment.
  S <- x[1]
  E <- x[2]
  I <- x[3]
  R <- x[4]
  
  # Get the model parameters.
  beta <- parameters$beta
  rho <- parameters$rho
  gamma <- parameters$gamma
  mu <- parameters$mu
  
  # Calculate the total number of individuals in the system.
  # This is just all the compartments added together.
  N <- S + E + I + R
  
  # Calculate the derivatives (the system of ODEs).
  dS <- mu*N - beta*I*S/N - mu*S
  dE <- beta*S*I/N - (mu + rho)*E
  dI <- rho*E - (mu + gamma)*I
  dR <- gamma*I - mu*R
  
  # Return the derivatives inside a list.
  derivatives <- list(c(dS, dE, dI, dR))
  return(derivatives)
}

```

```{r}
# The initial parameters: how many people are in each compartment at the start of
# the model?
init <- c(S = 997, E = 2, I = 1, R = 0.0)

# The model parameters: what is the transmission rate (beta), progression rate (rho) and recovery rate (gamma).
pars <- list(beta = 1.7, rho = 0.2, gamma = 0.1, mu = 0.0002)

# The solution times: the sequence of times for which to compute the solution.
time <- seq(from = 0, to = 100, by = 1)
```

```{r}
# lsoda is a function with inputs: ODE system (as a function), initial conditions,
# parameters and times to compute the solution. The solution for the SIR model is
# stored in 'out'.
out <- lsoda(func = modified.SEIR.model, y = init, parms = pars, times = time)

# Convert the solution into a data.frame. This makes plotting easier.
out <- as.data.frame(out)
head(out)
```
```{r}
myplotter(out = out)
```


```{r}
# view the last days of the model
tail(round(out))
```
## Exercise 2
Add temporary immunity of 9 days

```{r}
temp.modified.SEIR.model <- function(t, x, parameters){ 
  # Get the current number of individuals in each compartment.
  S <- x[1]
  E <- x[2]
  I <- x[3]
  R <- x[4]
  
  # Get the model parameters.
  beta <- parameters$beta
  rho <- parameters$rho
  gamma <- parameters$gamma
  mu <- parameters$mu
  w <- parameters$w
  
  # Calculate the total number of individuals in the system.
  # This is just all the compartments added together.
  N <- S + E + I + R
  
  # Calculate the derivatives (the system of ODEs).
  dS <- mu + w*R - beta*I*S/N - mu*S
  dE <- beta*S*I/N - (mu + rho)*E
  dI <- rho*E - (mu + gamma)*I
  dR <- gamma*I - w*R - mu*R
  
  # Return the derivatives inside a list.
  derivatives <- list(c(dS, dE, dI, dR))
  return(derivatives)
}
```

```{r}
# The initial parameters: how many people are in each compartment at the start of
# the model?
init <- c(S = 997, E = 2, I = 1, R = 0.0)

# The model parameters: what is the transmission rate (beta), progression rate (rho) and recovery rate (gamma).
pars <- list(beta = 1.7, rho = 0.2, gamma = 0.1, mu = 0.0002, w = 1/9)

# The solution times: the sequence of times for which to compute the solution.
time <- seq(from = 0, to = 100, by = 1)
```

```{r}
# lsoda is a function with inputs: ODE system (as a function), initial conditions,
# parameters and times to compute the solution. The solution for the SIR model is
# stored in 'out'.
out <- lsoda(func = temp.modified.SEIR.model, y = init, parms = pars, times = time)

# Convert the solution into a data.frame. This makes plotting easier.
out <- as.data.frame(out)
head(out)
```
```{r}
myplotter(out = out)
```


```{r}
# view the last days of the model
tail(round(out))
```


## Exercise 3
With the vital dynamics mantained, add a vaccination rate of 0.0005 with permanent immunity after both vaccination and recovery.

```{r}
vac.modified.SEIR.model <- function(t, x, parameters){ 
  # Get the current number of individuals in each compartment.
  S <- x[1]
  E <- x[2]
  I <- x[3]
  R <- x[4]
  V <- x[5]
  
  # Get the model parameters.
  beta <- parameters$beta
  rho <- parameters$rho
  gamma <- parameters$gamma
  mu <- parameters$mu
  w <- parameters$w
  v <- parameters$v
  
  # Calculate the total number of individuals in the system.
  # This is just all the compartments added together.
  N <- S + E + I + R + V
  
  # Calculate the derivatives (the system of ODEs).
  dS <- mu - beta*I*S/N - mu*S - v*S
  dE <- beta*S*I/N - (mu + rho)*E
  dI <- rho*E - (mu + gamma)*I
  dR <- gamma*I - mu*R
  dV <- v*S - mu*V
  
  # Return the derivatives inside a list.
  derivatives <- list(c(dS, dE, dI, dR, dV))
  return(derivatives)
}
```

```{r}
# The initial parameters: how many people are in each compartment at the start of
# the model?
init <- c(S = 997, E = 2, I = 1, R = 0.0, V = 0.0)

# The model parameters: what is the transmission rate (beta), progression rate (rho) and recovery rate (gamma).
pars <- list(beta = 1.7, rho = 0.2, gamma = 0.1, mu = 0.0002, w = 1/9, v = 0.0005)

# The solution times: the sequence of times for which to compute the solution.
time <- seq(from = 0, to = 100, by = 1)
```

```{r}
# lsoda is a function with inputs: ODE system (as a function), initial conditions,
# parameters and times to compute the solution. The solution for the SIR model is
# stored in 'out'.
out <- lsoda(func = vac.modified.SEIR.model, y = init, parms = pars, times = time)

# Convert the solution into a data.frame. This makes plotting easier.
out <- as.data.frame(out)
head(out)
```
```{r}
# First plot time (x-axis) against the number of susceptible (y-axis).
plot(out$time, out$S,
     type = 'l', col = 'green', ylab = "Number of individuals in each compartment",
     xlab = "Time (days)", lwd = 3, ylim = c(0, 1000), xlim = c(0,103))

# Next, we can add lines for time against infecteds, and time against recovered/removed.
lines(out$time, out$E, col = 'blue', lwd = 3)
lines(out$time, out$I, col = 'red', lwd = 3)
lines(out$time, out$R, col = 'grey', lwd = 3)
lines(out$time, out$V, col = 'black', lwd = 3)

# Finally, we can add a legend so we know what lines correspond to what compartment.
legend("topright", legend = c("S","E","I","R","V"), col = c("green","blue","red","grey","black"), lwd = 2)
```

```{r}
# view the last days of the model
tail(round(out))
```

i) What is the Ro of the final model and what does it imply?

Ro = beta/(gamma + mu)
Ro = 1.7/(0.1 + 0.0002)
Ro = 16.96607
Ro = 17

The Ro of 17 implies that each infectious individual is capable of infecting 17 other susceptible individuals.


ii) What proportion of individuals would have to be vaccinated to attain herd immunity for the final model?
Proportion, P of individuals to be vaccinated to attain herd immunity for the final model is as;

P = 1 - (1/Ro)
P = 1 - (1/17)
P = 0.9411765


